---
title: "Introduction to Single Cell RNAseq Part 3"
author: "UCD Bioinformatics Core"
output:
    html_document:
      keep_md: TRUE
---


Last Updated: March 24 2021, 11am

# Part 3: Batch Correction Excercise


## Load libraries
```{r libraries, warning=FALSE,error=FALSE,message=FALSE}
library(Seurat)
```

## Load the Seurat object from the prior excercise, and **create a batch effect**

```{r load_rdata, warning=FALSE,error=FALSE,message=FALSE}
load(file="pre_sample_corrected.RData")
experiment.aggregate
experiment.test <- experiment.aggregate
VariableFeatures(experiment.test) <- rownames(experiment.test)
set.seed(12345)


samplename = experiment.aggregate$orig.ident
rand.cells <- sample(1:ncol(experiment.test), 2000,replace = F)
batchid = rep("Example_Batch1",length(samplename))
batchid[rand.cells] = "Example_Batch2"
names(batchid) = colnames(experiment.aggregate)

experiment.test <- AddMetaData(
  object = experiment.test,
  metadata = batchid,
  col.name = "example_batchid")

table(experiment.test$example_batchid)

mat <- as.matrix(GetAssayData(experiment.test, slot="data"))

rand.genes <- sample(VariableFeatures(experiment.test), 500,replace = F)

mat[rand.genes,experiment.test$example_batchid=="Example_Batch2"] <- mat[rand.genes,experiment.test$example_batchid=="Example_Batch2"] + 0.22
experiment.test = SetAssayData(experiment.test, slot="data", new.data= mat )
rm(mat)
```

## Exploring Batch effects, none, Seurat [vars.to.regress]

First lets view the data without any corrections

## PCA in prep for tSNE

ScaleData - Scales and centers genes in the dataset.
```{r scaledata_help, warning=FALSE,error=FALSE,message=FALSE, eval=FALSE}
?ScaleData
```


```{r scale_data_no_adjustment, warning=FALSE,error=FALSE,message=FALSE}
experiment.test.noc <- ScaleData(object = experiment.test)
```

### Run PCA
```{r pca_none, warning=FALSE,error=FALSE,message=FALSE, fig.align="center", fig.caption="Uncorrected PCA plot"}
experiment.test.noc <- RunPCA(object = experiment.test.noc)
DimPlot(object = experiment.test.noc, group.by = "example_batchid", reduction = "pca")
DimPlot(object = experiment.test.noc, group.by = "example_batchid", dims = c(2,3), reduction = "pca")
```

PCA Elbow plot to determine how many principal components to use in downstream analyses.  Components after the "elbow" in the plot generally explain little additional variability in the data.

```{r pca_elbow, warning=FALSE,error=FALSE,message=FALSE}
ElbowPlot(experiment.test.noc)
```

We use 10 components in downstream analyses. Using more components more closely approximates the full data set but increases run time.

### TSNE Plot
```{r tsne, warning=FALSE,error=FALSE,message=FALSE, fig.align="center", fig.caption="Uncorrected TSNE plot"}
pcs.use <- 10
experiment.test.noc <- RunTSNE(object = experiment.test.noc, dims = 1:pcs.use)
DimPlot(object = experiment.test.noc,  group.by = "example_batchid")
```

## Correct for sample to sample differences (seurat)

Use vars.to.regress to correct for the sample to sample differences and percent mitochondria
```{r scaledata_regress, warning=FALSE,error=FALSE,message=FALSE, fig.align="center", fig.caption="Corrected PCA Plot"}
experiment.test.regress <- ScaleData(object = experiment.test,
                    vars.to.regress = c("example_batchid"), model.use = "linear")

experiment.test.regress <- RunPCA(object =experiment.test.regress,features=rownames(experiment.test.noc))

DimPlot(object = experiment.test.regress, group.by = "example_batchid", reduction = "pca")
```

### Corrected TSNE Plot
```{r tsne_2, warning=FALSE,error=FALSE,message=FALSE, fig.align="center", fig.caption="Corrected TSNE Plot"}
experiment.test.regress <- RunTSNE(object = experiment.test.regress, dims.use = 1:50)
DimPlot(object = experiment.test.regress, group.by = "example_batchid", reduction = "tsne")
```

#### Question(s)

1. Try a couple of PCA cutoffs (low and high) and compare the TSNE plots from the different methods.  Do they look meaningfully different?

### Excercise

Now go back to the original data without having been modified and see if a "batch effect" exists between the two actual batches

## Get the next Rmd file
```{r get_next_rmd, warning=FALSE,error=FALSE,message=FALSE, eval=FALSE}
download.file("https://raw.githubusercontent.com/ucdavis-bioinformatics-training/2021-August-Single-Cell-RNA-Seq-Analysis/master/data_analysis/scRNA_Workshop-PART4.Rmd", "scRNA_Workshop-PART4.Rmd")
```

## Session Information
```{r sessioninfo, warning=FALSE,error=FALSE,message=FALSE}
sessionInfo()
```
